Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-09-19 revision/r2/response_r2.tex

First, we would like to answer about the time it may take for this type of workflow to become more common in research spheres. A large portion of scientific results today are investigated and written by non-tenured scientists. We have examples around us of graduate students who are likely more inclined than their supervisors to spend time developing carefully checked models and following workflow steps that give them more confidence in their results. While their supervisors may not push them to spend more time doing so, they will not prevent them from doing it either. Therefore, we believe that this paper---among many others---could actually help spread the use of robust modeling workflows (faster than it seems). Although the road between being a graduate student and obtaining a tenure-track position is long and uncertain, those graduate students are soon to become reviewers and submit grants (if they decide to pursue in academia), and interact with decision-makers, at least at a local scale (and even for those who will quit academia).

\begin{mybox}
\emph{The above scenario (which can be applied to global analyses with or without workflows) is not
to deflect from the authors' main point being made that “that an improved workflow that
required data simulation and retrodictive checks would lead to larger model advances and a
greater recognition of uncertainty—thus highlighting likely consistency in estimates across
models—that could better aid policy”. The “could better aid policy” part is simply too vague
and could betray, in my opinion, that the proposed “universal framework” is implicitly
embracing the linear model of science (Fernándes 2016) of 'truth speaks to power'. In other
words, bad policy decisions are assumed to mainly result from a deficit of knowledge so that
all that is needed would be better knowledge to naturally lead to good policy decision. I find
this less and less convincing given that we’ve never had so much data or knowledge than now.
We know what are the main drivers of biodiversity loss (e.g. Maxwell et al. 2016). We are
literally awash with models (just think about the number of Species Distribution Models being
published nowadays). Yet biodiversity, on average, does not fare any better.}  
\end{mybox}

We agree that the main obstacles---at least at the global scale---for effective climate and biodiversity policies are not a lack of knowledge. However, coming back to your previous point (``\emph{Local, national or regional scientists involved in policy are then contacted to explain any discrepancies}''), we believe a different interpretation of the usefulness of the workflow for policy is possible. You noted above that scientists and decision-maskers can spend time discussing discrepancies with a high-profile global study---rather than on better decision-making or further research. We believe that a more systematic use of this kind of workflow would reduce such discrepancies (between different studies, estimates) and methodological controversies (a point we develop in the manuscript). So in this way, an improved workflow could aid policy.

@@

\begin{mybox}
\emph{While I agree with the authors that the workflow will help research spheres in disciplining
modelling, I disagree with the policy implications. As highlighted in the box by the authors,
“[m]odel development is the central step of the process-based workflow, typically requiring
several years, yet it often remains opaque for anyone who has not worked the model itself.”
This is not a small issue.}  
\end{mybox}

We wanted to highlight that process-based model development can sometimes be very long (with or without this workflow), and thus that it would be better to follow coherent steps in model development and checking---given the amount of time process-based model development takes anyway. But to be clear, for most applications, applying the workflow do not require years.

\begin{mybox}
\emph{For example, in the International Whaling Commission, established in
1946, many scientists are now retiring without new ones stepping in their shoes at the Scientific
Committee (SC). The IWC SC pioneered some work on simulations (e.g. De la Mare 1986) and
devised complex population models that were heavily tested to assess whether they would be
fit-for-purposes. It is ironic that the simulation-savvy and model-based procedures
encapsulated in the Revised Management Procedure that was developed in the late 1980s and
early 1990s was never used in practice (for policy reasons, not scientific ones, to make a long
story short).}  
\end{mybox}

We are not familiar with this example. Even without considering policy reasons, we do feel however that science sometimes goes in circles---and that great improvements in computational tools since 90s have not necessarily led to more robust models.

@@

...

\begin{mybox}
\emph{- ML enthusiasm (lines 298-303): the point is granted for prediction but again evidencing a
trend for policy is often required for further action. I fail to see how ML, if focused on
prediction, is going to help. A trend is a useful fiction that is easily understood in policy
fora where interpretability still has a huge premium. And a trend is very much
interpretable. I think the authors may be going overboard in their claims if, again, their
proposal is to have policy-relevance and not only research-relevance (a worthy goal of its
own).}  
\end{mybox}

Sorry it was not clear. Our point here was not to share any enthusiasm for machine learning, but rather to highlight the need for an even more careful and coherent workflow to build interpretable models (i.e. avoiding  ``complexity trap'' and build models with ``clear mechanistic drivers''). We tried to make this point more explicit  (\href{file:forecastflows_r2\#lintarget:clearerML}{line \ref*{lin:clearerML}}).

@@

\begin{mybox}
\emph{- Robustness is great. Policy loves the word but too rarely elaborates on it. Thus, on line 62,
these methods should be robust against what more precisely?}  
\end{mybox}

...

\begin{mybox}
\emph{- typo on line 138: “These feedback loops are applied currently in only in Bayesian
frameworks (to our knowledge)...”}
\end{mybox}

Thank you for catching this! We fixed this typo.

\begin{mybox}
\emph{Last, I would like to re-iterate that I view this manuscript as a very useful contribution for the
research spheres in ecology. It can be published as is for this audience. I do not dispute any of
the claims made with respect to research, namely that disciplining model building to the
rationalizing force of workflows is a good thing. I dispute what I think are shortcuts and could
be perceived as a naïve outlook on the policy process. I do not assume that the authors
subscribe to the linear model of science but a too quick reading of their opinion piece may
suggest otherwise, especially the introduction when it refers to policy. For example, if lines 32-
41 are an accurate description of the research sphere, the policy sphere is then very rational in
keeping an instrumental outlook to trend detection and favouring simple methods until the
research sphere gets its workflows fully operational. The path from an improved workflow in
research to better policy decisions remains to be charted: scientific workflows are likely a
necessary step, yet one that may not be sufficient.}
\end{mybox}

### 2025-09-19 revision/r2/response_r2.tex

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\RequirePackage[labelfont={bf,sf},%
                font={small, sf}]{caption}

\externaldocument{forecastflows_r2}
\newcommand{\lref}[1]{}
% (\href{file:forecastflows_r1\#lintarget:SOMELABEL}{line \ref*{lin:SOMELABEL}})

% \DeclareEmphSequence{\itshape\color{darkgrey}}
\usepackage[most]{tcolorbox}
% \newtcolorbox{mybox}[1][]{%
%     blanker, 
%     left skip=0em,
%     left=1em,
%     borderline west={1pt}{2pt}{darkgrey},
%     breakable,
%     #1}
\newtcolorbox{mybox}[1][]{%
    blanker, 
    left skip=0em,
    left=0em,
    borderline west={0pt}{0pt}{white},
    breakable,
    coltext=darkgrey
    #1}

@@

We are happy to hear that the three reviewers recognized the value of our work and that the manuscript is ready for publication. We have worked to fix the last small points raised by the reviewers, and we would like to thank them again for their careful reading. Furthermore, we also particularly appreciate the really interesting perspective on the policy implications from M. Authier, and...

{\bf Response to reviewers' comments:}

{\bf Reviewer 1}

\begin{mybox}
\emph{The authors have addressed all of my concerns and I believe the MS is
(nearly) ready for publication. I suggest one more pass for copy editing
to fix small mistakes like inconistencies for example in ``figure"
``Figure" and ``e.g." ``e.g.,".}
\end{mybox}

@@

{\bf Reviewer 2}

\begin{mybox}
\emph{All of my major comments were sufficiently addressed in this revision of
the manuscript. I appreciate that the supplement is now referenced in
the main document (I missed it on my last round of review). I have a few
editorial comments for the main text and a few minor comments about the
supplement, but I feel this manuscript is almost ready for publication.}
\end{mybox}

We are glad that the reviewer found the manuscript has improved.
Thank you for this last feedback.

\begin{mybox}
\emph{\textbf{Manuscript Minor comments:}\\
Abstract: potential typo in second-to-last sentence. Should ``where
forecasting a natural output" be ``where forecasting is a natural output"?}
\end{mybox}
% vvdm18sept: is really a typo? I thought it was just a way to say it. Lizzie?
...

@@

\begin{mybox}
\emph{Line 39: should "ignore" be "ignores"}
\end{mybox}

Thanks for catching this, it is fixed now.

\begin{mybox}
\emph{Line 51: ``hypotheses" or ``a hypothesis" rather than ``hypothesis"}
\end{mybox}

@@

Corrected.

\begin{mybox}
\emph{Line 138: there appears to be an extra ``in" in ``in only in"}
\end{mybox}

@@

Corrected.

\begin{mybox}
\emph{Line 215: comma after parameters}
\end{mybox}

Added.

\begin{mybox}
\emph{A couple typos in Box e.g., ``calibrating submodes separately allows to
avoid" missing "us"?}
\end{mybox}

@@

\begin{mybox}
\emph{Stan model not included so could not be assessed (model1\_nc2.stan), I
recommend sharing for transparency.}
\end{mybox}

Sorry we forgot to share the Stan code. We added the code in the workflow example.

\begin{mybox}
\emph{Figure 15, unclear what the posterior predictive distribution is of,
this detail could be added for transparency. Could also add many
posterior predictive datasets to the same plot and then overlay the
observed data to facilitate easier comparisons between y-tilde and y-obs.}
\end{mybox}

% Hmm I'm not sure to understand what he wants?
We now specify that the first posterior predictive distribution we show corresponds to the predicted counts across all species and populations.

@@

{\bf Reviewer 3}

\begin{mybox}
\emph{First, I would like to congratulate the authors for a very clear and well written manuscript. The main thesis of the authors is to outline modelling workflows and illustrate in broad brush strikes how to bridge the current gap in ecological forecasting with simulations of synthetic data among the several steps of model building (the pre-data/post-model third corner in their Fig. 1) to improve confidence in modelling, inference and prediction. This is an important contribution, one that echoes many early calls (e.g. Thomas 1997): our community in ecology, despite its beefing up with respect to quantitative skills, remains fragmented and disjoint. I thus wholeheartedly endorse the main thesis of the authors and agree with it about research. There remains one concern that I would like to discuss with the authors, and possibly hear back from them although it is somewhat minor with respect to research but more problematic with respect to policy. It should not deflect from publication of the article in Philosophical Transactions of the Royal Society A.}
\end{mybox}

We are glad that the reviewer found our manuscript well-written, and that he endorses our main messages. We are particularly grateful for his detailed comments, with much broader perspectives than just our manuscript. We believe that they deserve longer discussions (and could constitute in themselves another paper!), so we will reach out directly to M. Authier to share our detailed perspective on this.

@@

This is a very interesting question. However, we believe that the previous example of Callaghan et al. illustrates how bridging the differential goals of understanding and forecasting (or extrapolating to poorly observed species) can actually have important implications for policy. Similarly, it does not seem entirely accurate to think that this type of workflow should have been applied earlier (Revised Management Procedure of IWC) while also considering that it might be now 'too late' to highlight its importance (because there is no more time to lose to adress biodiversity loss?). We think this issue is linked to other challenges we encouter as scientists today: is there a risk in better quantifying and communicating the uncertainties of our results, if these uncertainties lead decision-makers to avoid making decisions? Or, by being the most transparent about our uncertain results (and spending more time applying consistent workflows), do we rather serve not only research but also the policy sphere?

\begin{mybox}
\emph{I would love to actually read the authors’ thoughts here and ideally be proven wrong that an
improved workflow would be enough for enabling policy decisions whose beneficial impacts
would be commensurate with the current magnitude of biodiversity loss7. In particular, I
would be interested in a better articulation of how they see global studies feed-backing on
policy which is usually more local, national or regional.}  
\end{mybox}

...

\begin{mybox}
\emph{My remaining points are more concise.
- Terminology. If the authors are also aiming at the policy sphere, the word bias (e.g. lines
13, 134) should be wielded with extreme caution as it is easily weaponized. In policy
circles, the term bias is, in my experience, the kiss of death as it is often interpreted as
lopsided evidence. The authors are discussing preferential sampling with respect to space,
time or taxonomy and its consequences in modelling results. Preferential sampling is more
descriptive than bias, which also has a prescriptive meaning, esp. in policy8. Also, a well-
designed survey at some spatial scale may become an instance of preferential geographic
sampling when collated with other surveys and data. This sort of data collation is often the
case with global trend analyses that the authors seem to have in mind.}  
\end{mybox}

\begin{mybox}
\emph{Also, taxonomic biases are bound to happen in global analyses. How do we redress those? That is, what can
we say about data-poor species? How a richly-parametrized workflow calibrated on data
collected on a few very well-studied species is going to help the majority of data-poor
species? Taxonomic gaps have been identified without workflows and how to address
those gaps point to monitoring and the need of appropriate governance to ensure long-
term monitoring and inclusion of data-poor species. Research alone may not provide
enough incentive or impetus for inclusion of data-poor species, especially if what is
needed for policy about them is descriptive and routine science.}  
\end{mybox}

@@

\begin{mybox}
\emph{- ML enthusiasm (lines 298-303): the point is granted for prediction but again evidencing a
trend for policy is often required for further action. I fail to see how ML, if focused on
prediction, is going to help. A trend is a useful fiction that is easily understood in policy
fora where interpretability still has a huge premium. And a trend is very much
interpretable. I think the authors may be going overboard in their claims if, again, their
proposal is to have policy-relevance and not only research-relevance (a worthy goal of its
own).}  
\end{mybox}

Sorry it was not clear. Our point here was not to share any enthusiasm for machine learning, but rather to highlight the need for an even more careful and coherent workflow to build interpretable models (i.e. avoiding  ``complexity trap'' and build models with ``clear mechanistic drivers''). We tried to make this point more explicit  (\href{file:forecastflows_r2\#lintarget:clearerML}{line \ref*{lin:clearerML}}).

\begin{mybox}
\emph{- Robustness is great. Policy loves the word but too rarely elaborates on it. Thus, on line 62,
these methods should be robust against what more precisely?}  
\end{mybox}

We have made this point clearer (\href{file:forecastflows_r2\#lintarget:robust}{line \ref*{lin:robust}}):
\begin{quote}
``This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps.''
\end{quote}

@@

\begin{mybox}
\emph{- typo on line 138: “These feedback loops are applied currently in only in Bayesian
frameworks (to our knowledge)...”}
\end{mybox}

Thank you for catching this! We fixed this typo.

\begin{mybox}
\emph{Last, I would like to re-iterate that I view this manuscript as a very useful contribution for the
research spheres in ecology. It can be published as is for this audience. I do not dispute any of
the claims made with respect to research, namely that disciplining model building to the
rationalizing force of workflows is a good thing. I dispute what I think are shortcuts and could
be perceived as a naïve outlook on the policy process. I do not assume that the authors
subscribe to the linear model of science but a too quick reading of their opinion piece may
suggest otherwise, especially the introduction when it refers to policy. For example, if lines 32-
41 are an accurate description of the research sphere, the policy sphere is then very rational in
keeping an instrumental outlook to trend detection and favouring simple methods until the
research sphere gets its workflows fully operational. The path from an improved workflow in
research to better policy decisions remains to be charted: scientific workflows are likely a
necessary step, yet one that may not be sufficient.}
\end{mybox}

We would like to thank the reviewer once again for all these very interesting avenues for discussion---beyond the community of researchers (to whom this paper was primarily addressed).

### 2025-09-19 analyses/lpiworkflow.qmd
finalize mroe or less revisions

What is global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?

@@

Using a simple linear gets us started, but now we need to consider how we will model species and populations. Populations usually are nested in species in ecological models, so we might consider that to start. We also need to make sure our model yields one global estimate; for this we decide to model species trends over time as part of larger distribution. Our resulting model would then be:

$$\log(y_i) \sim \text{normal}(\mu_i, \sigma)$$

@@

$$\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}} \ , \ \sigma_{\alpha, \text{sp}}) \\
\mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1} \ , \ \sigma_{\alpha_1}) \\
\sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2} \ , \ \sigma_{\alpha_2}) $$

$$\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}} \ , \ \sigma_{\beta, \text{sp}}) \\
\mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1} \ , \ \sigma_{\beta_1}) \\
\sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2} \ , \ \sigma_{\beta_2}) $$

@@

@@

Here we start by visually comparing the posterior predictive distribution (red line) to observed data (black dashed line):
```{r}
#| fig-height: 3
#| fig-width: 3
#| echo: false

>>>>

What is the global trend over time in vertebrate species population numbers, both overall and how does it vary across species and populations?

@@

Using a simple linear gets us started, but now we need to consider how we will model species and populations. Populations usually are nested in species in ecological models, so we might consider that to start. In this first hierarchical model, we thus assume that populations are exchangeable---an assumption we might refine later. We also need to make sure our model yields one global estimate; for this we decide to model species trends over time as part of larger distribution. Our resulting model would then be:

$$\log(y_i) \sim \text{normal}(\mu_i, \ \sigma)$$

@@

$$\alpha_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\alpha, \text{sp}}, \ \sigma_{\alpha, \text{sp}})
\quad \mu_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_1}, \ \sigma_{\alpha_1}) 
\quad \sigma_{\alpha,\text{sp}} \sim \text{normal}(\mu_{\alpha_2}, \ \sigma_{\alpha_2}) $$

$$\beta_{\text{pop}[\text{sp}]}  \sim \text{normal}(\mu_{\beta, \text{sp}}, \ \sigma_{\beta, \text{sp}}) 
\quad \mu_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_1}, \ \sigma_{\beta_1}) 
\quad \sigma_{\beta,\text{sp}} \sim \text{normal}(\mu_{\beta_2}, \ \sigma_{\beta_2}) $$

(where all the $\sigma$ parameters are greater than zero)

@@

```{stan output.var = "compiled_model"}

data {
  int<lower=1> N;
  int<lower=1> Nsp;
  int<lower=1> Npop;
  vector[N] y;
  vector[N] year;
  array[Npop] int<lower=1,upper=Nsp> spid_perpop;
  array[N] int<lower=1,upper=Npop> popid;
}

transformed data{
  vector[N] logy = log(y);
}

parameters {
  real mu_alpha1; // overall species intercept
  real<lower=0> sigma_alpha1;

  real mu_alpha2;
  real<lower=0> sigma_alpha2;

  array[Nsp] real mu_alpha_sp; // species-level intercepts
  array[Nsp] real<lower=0> sig_species_alpha;

  array[Npop] real alpha_tilde_pop_sp; // non-centered population intercepts

  real mu_beta1; // overall species trend
  real<lower=0> sigma_beta1;

  real mu_beta2;
  real<lower=0> sigma_beta2;

  array[Nsp] real mu_beta_sp; // species-level slopes
  array[Nsp] real<lower=0> sig_species_beta;

  array[Npop] real beta_tilde_pop_sp; // non-centered population slopes

  real<lower=0> sigma;

}

transformed parameters{

  array[Npop] real alpha_pop_sp;
  array[Npop] real beta_pop_sp;

  for (p in 1:Npop) {

      alpha_pop_sp[p] = mu_alpha_sp[spid_perpop[p]] 
      + sig_species_alpha[spid_perpop[p]] *  alpha_tilde_pop_sp[p];
      beta_pop_sp[p] = mu_beta_sp[spid_perpop[p]] 
      + sig_species_beta[spid_perpop[p]] *  beta_tilde_pop_sp[p];

  }

}

model {

  vector[N] mu;

  mu_alpha_sp ~ normal(mu_alpha1, sigma_alpha1);
  sig_species_alpha ~ normal(mu_alpha2, sigma_alpha2);

  mu_beta_sp ~ normal(mu_beta1, sigma_beta1);
  sig_species_beta ~ normal(mu_beta2, sigma_beta2);

  for (p in 1:Npop) {

    alpha_tilde_pop_sp[p] ~ normal(0,1);
    beta_tilde_pop_sp[p] ~ normal(0,1);

  }

  for(i in 1:N){

    mu[i] = alpha_pop_sp[popid[i]] + beta_pop_sp[popid[i]] * (year[i] - 1980);
    logy[i] ~ normal(mu[i], sigma);

  }

  mu_alpha1 ~ normal(0, log(500) / 2.57); // -log(500) < mu_alpha1 < log(500)
  sigma_alpha1 ~ normal(0, log(100) / 2.57); // 0 < sigma_alpha1 < log(100)

  mu_alpha2 ~ normal(0, log(500) / 2.57); // -log(500) < mu_alpha1 < log(500)
  sigma_alpha2 ~ normal(0, log(50) / 2.57); // 0 < sigma_alpha1 < log(50)

  mu_beta1 ~ normal(0, 2 / 2.57); // -2 < mu_beta1 < 2
  sigma_beta1 ~ normal(0, 2 / 2.57); // 0 < sigma_beta1 < 2

  mu_beta2 ~ normal(0, 2 / 2.57); //-2 < mu_beta1 < 2
  sigma_beta2 ~ normal(0, 2 / 2.57); // 0 < sigma_beta1 < 2

  sigma ~ normal(0, log(100) / 2.57);

}

generated quantities {

  vector[N] logy_pred;

  for(i in 1:N){

    real mu = alpha_pop_sp[popid[i]] 
    + beta_pop_sp[popid[i]] * (year[i] - 1980);
    logy_pred[i] = normal_rng(mu, sigma);

  }

}

@@

Here we start by visually comparing the posterior predictive distribution of the predicted counts (across all species and populations, red line) to observed data (black dashed line):
```{r}
#| fig-height: 3
#| fig-width: 3
#| echo: false

### 2025-09-19 revision/r2/forecastflows_r2.tex
slowly continue revisions
finalize mroe or less revisions

\begin{figure}
	\centering
	\includegraphics{../../figures/figure_modeldataspaces_revised}
	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

This reality should drive researchers to use more robust and coherent methods. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.

>>>>

\begin{figure}
	\centering
	\includegraphics{../../figures/figure_modeldataspaces_revised2}
	\caption{Different trajectories of current (b-c) and proposed (d) workflows through model-data space during model development, from pre-model/pre-data to post-model/post-data quadrants.
	Panel A serves as a legend, showing where model development should begin (carefully thinking about the hypotheses to test) and when inference should be made, for either explanatory purpose (estimation) or forecasting (prediction).
	Panel B illustrates that, in current trend estimation approaches, most time is spent in the post-model/post-data quadrant, where models are directly fitted to empirical data---often with no prior evaluation of the model.
	Panel C illustrates the typical process-based model development where a lot of time is spent building and calibrating (fitting) separate submodels (represented by the separated ovals) without clear distinction between model building and model calibration and without evaluating the full model structure a priori. Additionally, the specific hypotheses the model is designed to test are rarely clearly stated.
	Panel D illustrates the workflow highlighted in this paper (see main text and Figure 2 for details). Ideally, it begins in the pre-model/pre-data quadrant, where the research question is defined. The next step moves to the post-model/pre-data quadrant, where a model is built---independently of the data. This is where we should spend a good deal of time evaluating model behavior using simulated data. Once we are satisfied with our model, we can move to the post-model/post-data quadrant, where the model is fitted to real data. Then, we perform retrodictive checks to compare model predictions to observations, which likely give us some feedback to refine our model.}
	\label{fig:modeldata}
\end{figure}

@@

\llabel{robust} This reality should drive researchers to use more coherent methods that remain robust for the intended inferential goals, even with imperfect data and knowledge gaps. But the current workflows combined with the challenges ecologists are facing---both in terms of data complexity and societal needs---instead may lead to persistent problems.
Trend estimation has focused mostly on fitting a model to empirical data (\llabel{quad1}i.e. post-model/post-data, Figure \ref{fig:modeldata}b)---without the checks (\llabel{quad2}post-model/pre-data) and likely feedbacks that often highlight uncertainty and related limitations in the model and/or data (Figure \ref{fig:modeldata}b). For forecasting, researchers have focused on making predictions with increasingly complex mechanistic models (Figure \ref{fig:modeldata}c), frequently obscuring the steps underlying model building and parameterization (\llabel{quad3}and without a clear distinction between pre- and post-data). Researchers often calibrate the different parts of these models separately, and fix some parameter values based on experiments and expert knowledge, to avoid problems when trying to fit the model as a whole.
Addressing these problems while accounting for the realities of working with ecological data requires a more comprehensive workflow.

@@

With rapid advances in machine learning, improving current methods to gain greater scientific insights to drive better forecasting seems increasingly important. \llabel{clearerML} A careful and coherent workflow is essential to preserve the interpretability of non–machine-learning models---their primary advantage. \llabel{MLvsPBM}Machine learning will likely surpass process-based models for forecasting accuracy, especially if the latter lack a robust estimation of their parameters and fall in a complexity trap---a trap with additional costs to interpretability of process-based models. Similarly, estimates of trends from empirical data using models without clear mechanistic drivers may soon offer fewer advantages over machine learning.
